---
title: Linear Mixed Models
date: 2026-08-29
---

## Model setup

$$\boldsymbol{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{Z}\boldsymbol{u} + \boldsymbol{\epsilon}$$

where 

- $\boldsymbol{y}$ is the vector of observations
- $\mathbf{X}$ is the design matrix for fixed effects
- $\boldsymbol{\beta}$ is the vector of fixed effects
- $\mathbf{Z}$ is the design matrix for random effects
- $\boldsymbol{u}$ is the vector of random effects
- $\boldsymbol{\epsilon}$ is the vector of residuals,

assuming

$$\begin{bmatrix} \boldsymbol{u} \\ \boldsymbol{\epsilon} \end{bmatrix} \sim N\left(\begin{bmatrix} \boldsymbol{0} \\ \boldsymbol{0} \end{bmatrix}, \begin{bmatrix} \mathbf{G}(\boldsymbol{\sigma}_g) & \boldsymbol{0} \\ \boldsymbol{0} & \mathbf{R}(\boldsymbol{\sigma}_r) \end{bmatrix}\right)$$

where $\mathbf{G}$ is the covariance matrix of random effects and $\mathbf{R}$ is the covariance matrix of residuals.

### Alternative model setup

$$\boldsymbol{y} = \mathbf{W}\boldsymbol{b} + \boldsymbol{\epsilon}$$

where $\mathbf{W} = \begin{bmatrix} \mathbf{X} & \mathbf{Z} \end{bmatrix}$ and $\boldsymbol{b} = \begin{bmatrix} \boldsymbol{\beta} \\ \boldsymbol{u} \end{bmatrix}$.


## Mixed model equation 

Also called Henderson's mixed model equation, the mixed model equation is given by

$$\underbrace{\begin{bmatrix} \mathbf{X}^\top \mathbf{R}^{-1} \mathbf{X} & \mathbf{X}^\top \mathbf{R}^{-1} \mathbf{Z} \\ \mathbf{Z}^\top \mathbf{R}^{-1} \mathbf{X} & \mathbf{Z}^\top \mathbf{R}^{-1} \mathbf{Z} + \mathbf{G}^{-1} \end{bmatrix}}_{\mathbf{C}} \begin{bmatrix} \boldsymbol{\beta} \\ \boldsymbol{u} \end{bmatrix} = \begin{bmatrix} \mathbf{X}^\top \mathbf{R}^{-1} \boldsymbol{y} \\ \mathbf{Z}^\top \mathbf{R}^{-1} \boldsymbol{y} \end{bmatrix}$$


- Best Linear Unbiased Estimator (BLUE): $\hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{V}^{-1} \boldsymbol{y}$ 
- Best Linear Unbiased Predictor (BLUP): $\tilde{\boldsymbol{u}} = \mathbf{G} \mathbf{Z}^\top \mathbf{V}^{-1} (\boldsymbol{y} - \mathbf{X}\hat{\boldsymbol{\beta}})$
- $\tilde{\boldsymbol{\epsilon}} = \mathbf{R}\mathbf{P}\boldsymbol{y}$

## Useful matrices

- $\mathbf{V} = \text{Var}(\boldsymbol{y}) = \mathbf{Z}\mathbf{G}\mathbf{Z}^\top + \mathbf{R}$

- $\mathbf{P} = \mathbf{V}^{-1} - \mathbf{V}^{-1} \mathbf{X} (\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{V}^{-1}$

- $\mathbf{C} = \begin{bmatrix} \mathbf{X}^\top \mathbf{R}^{-1} \mathbf{X} & \mathbf{X}^\top \mathbf{R}^{-1} \mathbf{Z} \\ \mathbf{Z}^\top \mathbf{R}^{-1} \mathbf{X} & \mathbf{Z}^\top \mathbf{R}^{-1} \mathbf{Z} + \mathbf{G}^{-1} \end{bmatrix}$

## Prediction error variance


- $\mathbf{C}^{XX} = \text{Var}(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}) =\text{Var}(\hat{\boldsymbol{\beta}}) = (\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{X})^{-1}$

- $\mathbf{C}^{ZZ} = \text{Var}(\tilde{\boldsymbol{u}} - \boldsymbol{u}) = \mathbf{G} - \mathbf{G} \mathbf{Z}^\top \mathbf{P} \mathbf{Z} \mathbf{G}$

- $\mathbf{C}^{XZ} = \text{Cov}(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}, \tilde{\boldsymbol{u}} - \boldsymbol{u}) = -(\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{V}^{-1} \mathbf{Z} \mathbf{G}$

- $\mathbf{C}^{ZX} = \text{Cov}(\tilde{\boldsymbol{u}} - \boldsymbol{u}, \hat{\boldsymbol{\beta}} - \boldsymbol{\beta}) = -\mathbf{G} \mathbf{Z}^\top \mathbf{V}^{-1} \mathbf{X} (\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{X})^{-1}$


$$\text{Var}\left(\begin{bmatrix} \hat{\boldsymbol{\beta}} - \boldsymbol{\beta} \\ \tilde{\boldsymbol{u}} - \boldsymbol{u} \end{bmatrix} \right) = \mathbf{C}^{-1} = \begin{bmatrix}\mathbf{C}^{XX} & \mathbf{C}^{XZ} \\ \mathbf{C}^{ZX} & \mathbf{C}^{ZZ} \end{bmatrix}$$


## Prediction 

$$\mathbf{D}\tilde{\boldsymbol{b}} = \begin{bmatrix} \mathbf{D}_{\beta} & \mathbf{D}_{u} \end{bmatrix} \begin{bmatrix} \hat{\boldsymbol{\beta}} \\ \tilde{\boldsymbol{u}} \end{bmatrix} = \mathbf{D}_{\beta}\hat{\boldsymbol{\beta}} + \mathbf{D}_{u}\tilde{\boldsymbol{u}}$$

$$\text{Var}(\mathbf{D}\tilde{\boldsymbol{b}} - \mathbf{D}\boldsymbol{b}) = \mathbf{D}\mathbf{C}^{-1}\mathbf{D}^\top = \mathbf{D}_\beta \mathbf{C}^{XX} \mathbf{D}_\beta^\top + \mathbf{D}_\beta \mathbf{C}^{XZ} \mathbf{D}_u^\top + \mathbf{D}_u \mathbf{C}^{ZX} \mathbf{D}_\beta^\top + \mathbf{D}_u \mathbf{C}^{ZZ} \mathbf{D}_u^\top$$

## ASReml-R 

```{r}
#| warning: false
#| message: false
#| output: false
library(tidyverse)
library(asreml)
library(broom.asreml)
asreml.options(design = TRUE, 
               Cfixed = TRUE, 
               Csparse =~ Blocks/Wplots)
fit <- asreml(yield ~ Variety * Nitrogen,  
              random = ~ Blocks/Wplots, data = oats)

X <- model.matrix(~ Variety*Nitrogen, data = oats)
Z <- cbind(model.matrix(~ Blocks - 1, data = oats), model.matrix(~ Blocks:Wplots - 1, data = oats))
n <- nrow(X)
```

### Did the model converge?

```{r}
fit$converge
```

### Residual log-likelihood 

```{r}
fit$loglik
```

### Residual variance 

If $\mathbf{R} = \sigma^2 \mathbf{I}$, then the residual variance is given by

```{r}
fit$sigma2
```

```{r}
R <- fit$sigma2 * diag(n)
```


### Variance parameters

```{r}
fit$vparameters
```

```{r}
G <- fit$sigma2 * Matrix::bdiag(fit$vparameters[1] * diag(6), fit$vparameters[2] * diag(18))
```

### Coefficients

```{r}
fit$coefficients
```



```{r}
V <- Z %*% G %*% t(Z) + R
Vinv <- solve(V)
P <- Vinv - Vinv %*% X %*% solve(t(X) %*% Vinv %*% X) %*% t(X) %*% Vinv


(beta_hat <- solve(t(X) %*% Vinv %*% X) %*% t(X) %*% Vinv %*% oats$yield)
(u_tilde <- G %*% t(Z) %*% Vinv %*% (oats$yield - X %*% beta_hat))
```


### Variance-covariance matrix of coefficients

$\dfrac{1}{\hat{\sigma}^2}\text{diag}(\mathbf{C}^{-1})$

```{r}
fit$vcoeff
```

```{r}
C <- rbind(cbind(t(X) %*% solve(R) %*% X, t(X) %*% solve(R) %*% Z), cbind(t(Z) %*% solve(R) %*% X, t(Z) %*% solve(R) %*% Z + solve(G)))
Cinv <- solve(C)
diag(Cinv) / fit$sigma2
```

## Linear predictors

$\mathbf{X}\hat{\boldsymbol{\beta}} + \mathbf{Z}\tilde{\boldsymbol{u}}$

```{r}
fit$linear.predictors
```

```{r}
as.vector(X %*% beta_hat + Z %*% u_tilde)
```

## Prediction error variance

```{r}
image(fit$Cfixed)
```

```{r}
image(solve(t(X) %*% Vinv %*% X))
```

```{r}
fit$Csparse |> 
  asreml::sp2Matrix() |> 
  image()
```

```{r}
image(G - G %*% t(Z) %*% P %*% Z %*% G)
```

## Design matrices

$\mathbf{W} = \begin{bmatrix} \mathbf{X} & \mathbf{Z} \end{bmatrix}$

```{r}
image(fit$design)
```